---
title: "MLPA_KELS"
author: "Seongmin Park"
date: "2025-09-13"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: tango
---

# MLPA_KELS

## Data Preparation

Using data form KELS 2013 7th

```{r results = "hide", warning = FALSE, message = FALSE}

library(plyr)
library(tidyverse)
library(MplusAutomation)
library(haven)

path <- setwd("/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis")
```

Load your SPSS Data. Check your dataset already included in your designated path; if not, just specify each path of your data.

```{r}

stu_data <- read_spss("/Users/seongminpark/Desktop/KEMS/25-2/KELS/KELS2013_7차/1. L2Y7S_학생.sav")
sch_data <- read_spss("/Users/seongminpark/Desktop/KEMS/25-2/KELS/KELS2013_7차/5. L2Y7DB_학교DB.sav")
par_data <- read_spss("/Users/seongminpark/Desktop/KEMS/25-2/KELS/KELS2013_7차/3. L2Y7P_학부모.sav")
stuf_data <- read_spss("/Users/seongminpark/Desktop/KEMS/25-2/KELS/KELS2013_7차/2. L2Y7LT_학습자특성.sav")
```

Below is my personal data tidying function:

```{r}

# Overall Missing Num Check
missing_num_check <- function(data) {
  colSums(is.na(data))
}

# Listwise Deletion
listwise_del <- function(data) {
  data %>%
  filter(complete.cases(.))
}

# NO haven
NO_haven <- function(data) {
  data %>%
  mutate(across(where(is.labelled), ~ as.numeric(as.character(.))))
}

# NA = -999 function
NA_to_999 <- function(data) {
  data[is.na(data)] <- -999
  return(data)
}
```

## Variable Selection

Starts from checking the most adequate class number. All missing values are already assigned as NA.

At first, **you need to change the covariate lists for the better explaining model.**

```{r}

## Add covariates here

stu_cov_list <- c("L2GENDER")
sch_cov_list <- c("L2Y7_REG")
```

Below is the quick data tidying process. I just remove the row which includes all missings for LPA indicators; also with SCHID.

```{r}
# Variable Selection and excluding out the missing data
LPA_data <- stu_data %>% 
  select(L2SID, L2Y7_SCHID,
         starts_with("L2Y7S19"),
         all_of(stu_cov_list),
         all_of(sch_cov_list)) %>% 
  filter(!if_all(starts_with("L2Y7S19"), is.na)) %>% # n = 6,182
  filter(!is.na(L2Y7_SCHID)) # n = 6,148

# How many students and schools?
## School: n = 1,657, Student: n = 6,148. Mean cluster size = 3.71
nrow(LPA_data)

LPA_data %>%
  summarise(n_schools = n_distinct(L2Y7_SCHID))

# Remove haven label
LPA_data <- NO_haven(LPA_data) 

# NA to -999 for using data in mplus
LPA_data <- NA_to_999(LPA_data) 

# Write .dat file for mplus
write.table(LPA_data, "LPA_data.dat", sep = " ", col.names = F, row.names = F)
```

Data file needs to be cleaned more... since the list of variable is not yet organized.

------------------------------------------------------------------------

## Latent Profile Analysis using MplusAutomation

Too annoying to run all classes separately; Just use MplusAutomation :D\
Revise the following codes for automatic mplus input files.

For running the model in this analysis, you only need to change '**outputDirectory**' for your personal laptop, and **'VARIABLE', 'USEVARIABLES'**, depending on your selected covariates.

Check your outputdirectory and covariate list (colnames):

```{r}

# outputdirectory
getwd()

# full variable list in order
## VARIABLE: NAMES ARE
cat(gsub('"', '', colnames(LPA_data)), sep = " ")

# After deciding the number of classes, you need to add covariates syntax in variable section, which indicates the within and between level covariates. Just copy the code result of below syntax:
cat("within =", gsub('"', '', stu_cov_list), collapse = "")
cat("between =", gsub('"', '', sch_cov_list), collapse = "")
```

## 1. Single-level LPA

```         
[[init]]
iterators = classes;
classes = 1:6;
outputDirectory = "/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis";
filename = "LPA_Analysis_class_[[classes]].inp";
[[/init]]

TITLE: SLPA_Analysis_class_[[classes]]
DATA: FILE IS LPA_data.dat;

VARIABLE: NAMES ARE 
SID SCHID ex1-ex4 ac1-ac4
ST_SEX !within covariates
SC_REG; !between covariates
USEVARIABLES = ex1-ex4 ac1-ac4;
CLASSES = c([[classes]]);
MISSING = all(-999);

ANALYSIS: TYPE = MIXTURE;
STARTS = 100 20;

OUTPUT: TECH11 TECH14;
```

```{r eval = FALSE}

# simple copy & paste of above text
## [init] section is for building multiple mplus input syntax simultaneously.

SLPA_text <- '
[[init]]
iterators = classes;
classes = 1:6;
outputDirectory = "/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis";
filename = "SLPA_Analysis_class_[[classes]].inp";
[[/init]]

TITLE: SLPA_Analysis_class_[[classes]]
DATA: FILE IS LPA_data.dat;

VARIABLE: NAMES ARE 
SID SCHID ex1-ex4 ac1-ac4
ST_SEX !within covariates
SC_REG; !between covariates
USEVARIABLES = ex1-ex4 ac1-ac4;
CLASSES = c([[classes]]);
MISSING = all(-999);

ANALYSIS: TYPE = MIXTURE;
STARTS = 100 20;

OUTPUT: TECH11 TECH14;
'

writeLines(SLPA_text, "SLPA_text.txt")
```

Now, use `createModels()` to make multiple input files:

```{r eval = FALSE}

createModels("SLPA_text.txt")
```

Next, run all models in following directory: since I already attached the file directory by `setwd()`, simply running the following function still works:

```{r eval = FALSE}

# Use this method, since filefilter cannot recognize the regex ;
# SLPA_file_list <- list.files(".", pattern = "^SLPA.*.inp$")
# runModels(SLPA_file_list)

# Run simultaneously
runModels(filefilter = "SLPA")
```

And... readmodels.

```{r}

SLPA_output_list <- readModels(filefilter = "SLPA")
SLPA_summaryStats <- do.call("rbind.fill", sapply(SLPA_output_list, "[", "summaries"))
SLPA_summary <- SLPA_summaryStats %>% 
  select(Title, LL, AIC, BIC, Entropy, T11_LMR_PValue, BLRT_PValue) %>%
  mutate(AIC_diff = lag(AIC) - AIC) %>% 
  mutate(BIC_diff = lag(BIC) - BIC)

SLPA_summary
```

From the summary table, we could decide **class 3** (Largest BIC_diff) or **class 4** (LMR LRT is not significant at class 5)

```{r}

# Class 3

SLPA3_class_df <- SLPA_output_list[["SLPA_Analysis_class_3.out"]]$class_counts$posteriorProb

SLPA3_n_class_df <- SLPA_output_list[["SLPA_Analysis_class_3.out"]]$class_counts$mostLikely

SLPA3_means_df <- SLPA_output_list[["SLPA_Analysis_class_3.out"]]$parameters$unstandardized %>%
  filter(paramHeader == "Means", LatentClass != "Categorical.Latent.Variables") %>%
  select(LatentClass, param, est)

SLPA3_means_vars_df <- SLPA_output_list[["SLPA_Analysis_class_3.out"]]$parameters$unstandardized %>%
  filter(paramHeader == "Means" | paramHeader == "Variances", LatentClass != "Categorical.Latent.Variables") %>%
  select(LatentClass, paramHeader, param, est)

# Class Proportions (Posterior probabilities)
SLPA3_class_df

# Class Proportions (Most Likely probabilities)
SLPA3_n_class_df

# Class 3 Mean & Var

as.data.frame(matrix(SLPA3_means_vars_df$est, ncol = 8, byrow = TRUE,
                     dimnames = list(c("c1_mean", "c1_var", "c2_mean", "c2_var", "c3_mean", "c3_var"), c("ex1", "ex2", "ex3", "ex4", "ac1", "ac2", "ac3", "ac4"))))

# LPA plot
ggplot(SLPA3_means_df, aes(x = param, y = est, color = factor(LatentClass), group = LatentClass)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(title = "Latent Profile Means by Class (3)",
       x = "Indicator",
       y = "Estimated Mean",
       color = "Latent Class") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}

# Class 4

SLPA4_class_df <- SLPA_output_list[["SLPA_Analysis_class_4.out"]]$class_counts$posteriorProb

SLPA4_n_class_df <- SLPA_output_list[["SLPA_Analysis_class_4.out"]]$class_counts$mostLikely

SLPA4_means_df <- SLPA_output_list[["SLPA_Analysis_class_4.out"]]$parameters$unstandardized %>%
  filter(paramHeader == "Means", LatentClass != "Categorical.Latent.Variables") %>%
  select(LatentClass, param, est)

# Class Proportions (Most Likely probabilities)
SLPA4_n_class_df

ggplot(SLPA4_means_df, aes(x = param, y = est, color = factor(LatentClass), group = LatentClass)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(title = "Latent Profile Means by Class (4)",
       x = "Indicator",
       y = "Estimated Mean",
       color = "Latent Class") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since there is no graphical difference between Class 3 and Class 4, I will choose **Class 3,** for more parsimonious model.

------------------------------------------------------------------------

## 2. Multi-level LPA

```         
[[init]]
iterators = classes;
classes = 1:6;
outputDirectory = "/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis";
filename = "MLPA_Analysis_class_[[classes]].inp";
[[/init]]

TITLE: MLPA_Analysis_class_[[classes]]
DATA: FILE IS LPA_data.dat;

VARIABLE: NAMES ARE 
SID SCHID ex1-ex4 ac1-ac4
ST_SEX !within covariates
SC_REG; !between covariates
USEVARIABLES = ex1-ex4 ac1-ac4;
CLASSES = bc([[classes]]) wc(3); ! fix the best single-level LPA result
CLUSTER = SCHID;
WITHIN ARE ex1-ex4 ac1-ac4;
BETWEEN ARE bc;
MISSING = all(-999);

ANALYSIS: TYPE = MIXTURE TWOLEVEL;
STARTS = 100 20; ! need to increase the number

MODEL:
%WITHIN%
%OVERALL%

%BETWEEN%

%OVERALL%
wc ON bc;

MODEL wc:
%WITHIN%
%wc#1%
[ex1-ex4 ac1-ac4];
%wc#2%
[ex1-ex4 ac1-ac4];
%wc#3%
[ex1-ex4 ac1-ac4];
```

```{r eval = FALSE}

MLPA_text <- '
[[init]]
iterators = classes;
classes = 1:6;
outputDirectory = "/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis";
filename = "MLPA_Analysis_class_[[classes]].inp";
[[/init]]

TITLE: MLPA_Analysis_class_[[classes]]
DATA: FILE IS LPA_data.dat;

VARIABLE: NAMES ARE 
SID SCHID ex1-ex4 ac1-ac4
ST_SEX !within covariates
SC_REG; !between covariates
USEVARIABLES = ex1-ex4 ac1-ac4;
CLASSES = bc([[classes]]) wc(3); ! fix the best single-level LPA result
CLUSTER = SCHID;
WITHIN ARE ex1-ex4 ac1-ac4;
BETWEEN ARE bc;
MISSING = all(-999);

ANALYSIS: TYPE = MIXTURE TWOLEVEL;
STARTS = 100 20; ! need to increase the number

MODEL:
%WITHIN%
%OVERALL%

%BETWEEN%

%OVERALL%
wc ON bc;

MODEL wc:
%WITHIN%
%wc#1%
[ex1-ex4 ac1-ac4];
%wc#2%
[ex1-ex4 ac1-ac4];
%wc#3%
[ex1-ex4 ac1-ac4];
'

writeLines(MLPA_text, "MLPA_text.txt")
createModels("MLPA_text.txt")
runModels(filefilter = "MLPA")
```

```{r}

# Readmodels
MLPA_output_list <- readModels(filefilter = "MLPA")
MLPA_summaryStats <- do.call("rbind.fill", sapply(MLPA_output_list, "[", "summaries"))
MLPA_summary <- MLPA_summaryStats %>% 
  select(Title, LL, AIC, BIC, aBIC, Entropy) %>%
  mutate(AIC_diff = lag(AIC) - AIC) %>% 
  mutate(BIC_diff = lag(BIC) - BIC) 

MLPA_summary # class 2 shows the lowest BIC, but low entropy.
```

```{r}

# Class 2

MLPA_output_list[["MLPA_Analysis_class_2.out"]]$output

MLPA2_class_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$class_counts$posteriorProb

MLPA2_n_class_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$class_counts$mostLikely
MLPA2_pattern_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$class_counts$mostLikely.patterns

MLPA2_means_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$parameters$unstandardized %>%
  filter(paramHeader == "Means", LatentClass != "Categorical.Latent.Variables") %>%
  select(LatentClass, param, est)

MLPA2_means_vars_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$parameters$unstandardized %>%
  filter(paramHeader == "Means" | paramHeader == "Variances", LatentClass != "Categorical.Latent.Variables") %>%
  select(LatentClass, paramHeader, param, est)

MLPA2_means_vars_df <- as.data.frame(matrix(MLPA2_means_vars_df$est, ncol = 8, byrow = TRUE,
                     dimnames = list(c("c1_1mean", "c1_1var", "c1_2mean", "c1_2var", "c1_3mean", "c1_3var","c2_1mean", "c2_1var", "c2_2mean", "c2_2var", "c2_3mean", "c2_3var"), c("ex1", "ex2", "ex3", "ex4", "ac1", "ac2", "ac3", "ac4"))))

# Class Proportions (Most Likely probabilities)
MLPA2_n_class_df

# Class 2 Mean & Var
MLPA2_means_vars_df
```

```{r}

MLPA2_pattern_df <- MLPA_output_list[["MLPA_Analysis_class_2.out"]]$class_counts$mostLikely.patterns

MLPA2_pattern_df <- MLPA2_pattern_df %>% 
  dplyr::group_by(BC) %>% 
  dplyr::mutate(B_proportion = proportion / sum(proportion)) %>%
  ungroup()

# Plot (Within)

within_plot <- MLPA2_means_df %>% 
  filter(LatentClass == 1.1 | LatentClass == 1.2 | LatentClass == 1.3) %>% 
  mutate(LatentClass = case_when(
    LatentClass == 1.1 ~ "c1 (Medium)",
    LatentClass == 1.2 ~ "c2 (Low)",
    LatentClass == 1.3 ~ "c3 (High)"
  )
)

ggplot(within_plot, aes(x = param, y = est, color = factor(LatentClass), group = LatentClass)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(title = "Latent Profile Means by Class (3) - within",
       x = "Indicator",
       y = "Estimated Mean",
       color = "Latent Class") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot (Between)

df_labels <- MLPA2_pattern_df %>%
  dplyr::group_by(BC) %>%
  dplyr::arrange(BC, desc(WC)) %>%  # Arrange so labels don't overlap
  dplyr::mutate(
    ypos = cumsum(B_proportion) - 0.5 * B_proportion,
    label = scales::percent(B_proportion, accuracy = 1), 
    WC = case_when(
    WC == 1 ~ "c1 (Medium)",
    WC == 2 ~ "c2 (Low)",
    WC == 3 ~ "c3 (High)"
    )
    )
  
ggplot(df_labels, aes(x = factor(BC), y = B_proportion, fill = factor(WC))) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(y = ypos, label = label), color = "white", size = 4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(ylim = c(0, 1)) +  #  sets y-axis to 0–1 (0–100%)
  labs(
    title = "BC-WC plot",
    x = "Between-Class (BC)",
    y = "Within-BC Proportion (%)",
    fill = "Within-Class (WC)"
  ) +
  theme_minimal(base_size = 14)
```

Hence, I'll conclude nonparametric BC(2) WC(3) model. Base on the plot above, **school group 2** includes schools with more academic stress and exam stress, compared to **school group 1.**

------------------------------------------------------------------------

## 3. L1-L2 covariates

**`!!!!!!!!!!!!!We need to update this part to find the best fitting model!!!!!!!!!!!!!!`**

Bring the var_list again:

```{r}

# full variable list in order
## VARIABLE: NAMES ARE
cat(gsub('"', '', colnames(LPA_data)), sep = " ")

# After deciding the number of classes, you need to add covariates syntax in variable section, which indicates the within and between level covariates. Just copy the code result of below syntax:
cat("WITHIN ARE", gsub('"', '', stu_cov_list), ";", collapse = "")
cat("BETWEEN ARE bc", gsub('"', '', sch_cov_list), ";", collapse = "")
```

To fix the class mean, use this syntax and copy & paste:

```{r}

# Fix the observed mean and variance for each class
## c1
paste0(colnames(MLPA2_means_vars_df), "@", as.vector(MLPA2_means_vars_df[1,]), collapse = " ")
## c2
paste0(colnames(MLPA2_means_vars_df), "@", as.vector(MLPA2_means_vars_df[3,]), collapse = " ")
## c3
paste0(colnames(MLPA2_means_vars_df), "@", as.vector(MLPA2_means_vars_df[5,]), collapse = " ")
```

```{r eval = FALSE}

cov_text <- '
[[init]]
iterators = classes;
classes = 3;
outputDirectory = "/Users/seongminpark/Desktop/KEMS/25-2/KELS/Analysis";
filename = "Covariate_MLPA_bc2_wc[[classes]].inp";
[[/init]]

TITLE: Covariates_BC2_WC3_MLPA
DATA: FILE IS LPA_data.dat;

VARIABLE: NAMES ARE 
SID SCHID ex1-ex4 ac1-ac4
ST_SEX !within covariates need to be updated
SC_REG; !between covariates need to be updated
USEVARIABLES = ex1-ex4 ac1-ac4 ST_SEX SC_SEX;
CLASSES = bc(2) wc([[classes]]); ! fix the best single-level LPA result
CLUSTER = SCHID;
WITHIN ARE ex1-ex4 ac1-ac4 ST_SEX;
BETWEEN ARE bc SC_SEX;
MISSING = all(-999);

DEFINE:
SC_SEX = CLUSTER_MEAN(ST_SEX);

ANALYSIS: TYPE = MIXTURE TWOLEVEL;
STARTS = 0;
OPTSEED = 399671; ! Need to be improved

MODEL:
%WITHIN%
%OVERALL%
wc on ST_SEX;! Add L1 covariates

%BETWEEN%
%OVERALL%
wc on bc;
bc ON SC_SEX; ! Add L2 covariates


MODEL wc:
%WITHIN%
%wc#1%
[ex1@3.218 ex2@2.856 ex3@3.125 ex4@3.188 ac1@2.681 ac2@3.477 ac3@2.067 ac4@2.39];
%wc#2%
[ex1@1.554 ex2@1.371 ex3@1.419 ex4@1.532 ac1@1.723 ac2@1.703 ac3@1.28 ac4@1.513];
%wc#3%
[ex1@4.288 ex2@4.105 ex3@4.381 ex4@4.283 ac1@3.477 ac2@4.467 ac3@3.287 ac4@3.35];
'

writeLines(cov_text, "cov_text.txt")
createModels("cov_text.txt")
runModels(filefilter = "Covariate")
```

```{r}

Covariate_output <- readModels(filefilter = "Covariate")
Covariate_output$output
```
